Relation of severe COVID-19 in Scotland to transmission-related factors and risk conditions eligible for shielding support: REACT-SCOT case-control study

### 2021-08-29 transmission17Mar.Rmd
reinfections

# rmarkdown::render("shielding.Rmd", output_file="shielding.pdf")

@@

@@

Word counts: abstract 396 words, main text 5603 words.  
2 figures, 4 tables, 25 references

@@

# Abstract

* _Objectives_ -  To investigate:(1) the risk of severe  COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.  
* _Design_ - Matched case-control study (REACT-SCOT). 
* _Setting_ - Population of Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)`. 
* _Participants_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 and `r nrow(cc.all[CASE==0])` controls matched for age, sex and primary care practice, linked with all `r nrow(shielded.full)` individuals identified as eligible for shielding by Public Health Scotland.
* _Main outcome measure_ - Severe COVID-19, defined as cases that entered critical care or were  fatal.  
* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estcipv(table.listedgr3.severecases.nocare["Moderate risk condition", 3], table.listedgr3.severecases.nocare["Moderate risk condition", 4])` 
in those with moderate risk conditions and 
`r format.estcipv(table.listedgr3.severecases.nocare["Eligible for shielding", 3], table.listedgr3.severecases.nocare["Eligible for shielding", 4])` 
in those eligible for shielding.  The highest rate was in solid organ transplant recipients: rate ratio `r format.estcipv(table.shielded.severecases.nocare["Solid organ transplant", 3], table.shielded.severecases.nocare["Solid organ transplant", 4])`.  In both the shielded and  the general population, the risk of severe COVID-19 increased with the number of adults  but decreased with the number of school-age children in the household. Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estcipv(table.shielded.severecases.nocare["Recent hospital visit/stay", 3], table.shielded.severecases.nocare["Recent hospital visit/stay", 4])`  overall.   In a case-crossover analysis with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.  Among those eligible for shielding, the population attributable risk fraction (PARF) of severe cases for recent exposure to hospital was  `r round(100 * pc * (r -1) / r)`%.  In the general population the PARF for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.  
* _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.  For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts. Mitigating the impact of the epidemic requires control of nosocomial transmission.

\newpage
\linenumbers

# Summary

## What is already known on this topic?

* Individuals designated as "clinically extremely vulnerable" -- about 3% of the population -- have been advised to shield themselves during the COVID-19 epidemic

@@

## What this study adds

* Hospital-acquired infection has made a substantial contribution to the burden of COVID-19 in the population, accounting for more than half of severe cases in early December 2020.

* For vulnerable individuals to be shielded, the risk of hospital-acquired infection must be reduced and support must be provided for other adults in the household to co-isolate with the vulnerable individual.

@@

<!--- 
* don't blame HCW, there are problems 
* intel failure
--->

@@

# Methods
We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions". Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.  The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio as described below.

## Ascertainment of eligibility for shielding 
We used the dataset of those eligible for shielding that was created by PHS for the purposes of issuing letters.   This dataset comprised a list of  patient identifiers and the shielding condition and the date the shielding letter was issued. The list had been generated  by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries,  for evidence of a list of agreed diseases and conditions  and supplemented by data submitted  to PHS following direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].  The categories designated as eligible for shielding were as listed below [@health_protection_scotland_search_2020; @kanani_caring_2020]:

1. Solid organ transplant recipient
2. Cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
3. Severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
4. Rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle cell disease. 
5. People on immunosuppression therapies sufficient to increase risk of infection
6. Pregnant with heart disease
7. Additional conditions, including people on renal dialysis, those who had had a splenectomy  and others identified by clinicians as requiring shielding advice.

@@

## Transmission-related risk factors 
Since addresses in the CHI database have been mapped to  Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to refine the care home residence  assignation contained in the CHI database.  Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.  Linkage to occupational status for health-care workers and teachers was undertaken as described elsewhere [@shah_risk_2020; @mcallister_report_2020].

We used the Scottish Morbidity Records SMR01 (inpatients and day cases) and SMR00 (outpatient attendance) together with the RAPID database to derive variables encoding recent exposure to hospitals. We defined the variable "recent hospital exposure" as any hospital in-patient stay, day case attendance or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.  Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].  The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.  Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated. <!---The ECDPC states that "cases with symptom onset within 14 days of discharge from a healthcare facility (e.g. re-admission) may be considered as community-associated, probable or definite health-care-associated-COVID-19, or to have an indeterminate association" but does not specify how cases are to be assigned to these categories.--->

## Statistical methods
The analyses presented here focus on severe COVID-19, defined as the main outcome measure of  of the REACT-SCOT study at the design stage.  Ascertainment of this outcome does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is likely to differ between those eligible for shielding and those ineligible.  Care home residents were excluded from case-control analyses as shielding advice intended for individuals in private households would not have been relevant to care home residents.

@@

To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.  The population attributable risk fraction (PARF) of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.  where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].  Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.

@@

To test for causality of the association of recent exposure to hospital with severe COVID-19, we estimated rate ratios associated with recent hospital exposure (5 to 14 days before first testing positive) with less recent exposure (15 to 24 days before first testing positive) as reference category. This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases. It is a variant of the classic case-crossover design, which compares in cases only the frequencies of exposure in recent and less recent time windows. This can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000]. The conditional odds ratio can thus be estimated as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only. The analysis that we report is a refinement that takes advantage of the availability of a matched control group.  We contrast not the numbers of cases with recent exposure only and nonrecent exposure only, but the conditional odds of case status between those with recent exposure only and those with less recent exposure only. This controls for any difference in the population frequencies of exposure or disease incidence between recent and less recent time windows.

@@

In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r format.estcipv(table.rapid.coeffs["rapid.recent", 3], table.rapid.coeffs["rapid.recent", 4])`]; the  rate ratios associated with day case exposure [`r format.estcipv(table.daycase.coeffs["daycase.recent", 3], table.daycase.coeffs["daycase.recent", 4])`] or outpatient exposure  [`r format.estcipv(table.opd.coeffs["opd.recent", 3], table.opd.coeffs["opd.recent", 4])`]  were  much lower.   To test if the association of severe COVID-19 with recent inpatient exposure was likely to simply reflect  confounding by time invariant comorbidity, the association was examined by time window as described in the Methods section. Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.   
<!--- tabulate underlying cause of death on death certs of fatal cases
paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(hosp.recent, covid_ucod)))
with(cc.severe[CASE==1], table(hosp.recent, inhosp))
paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(occup, covid_ucod)))
--->
The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.  From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.  As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital acquired infection using the ECPDC classification.   The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.

@@

### Risk to those eligible for shielding
In  comparison with those with no risk condition, the highest rate ratio for severe COVID-19 was the `r round(as.numeric(gsub(" .*", "", table.shielded.severecases.nocare["Solid organ transplant", 3])))`-fold rate ratio in solid organ transplant recipients. For other conditions deemed eligible for shielding, the rate ratios were between `r round(min(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[4:8, 3]))))` and `r round(max(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[4:8, 3]))))`, compared to the rate ratio of about `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[2, 3])))` associated with conditions designated as moderate risk, which include heart disease, diabetes, chronic kidney disease and disabling neurological conditions.  The numbers of pregnant women with heart disease were too small for the risk in this group to be estimated. Other studies of outcome of COVID-19 in solid organ transplant recipients and other individuals using immunosuppressants have been based only on patients admitted to hospital [@avery_inpatient_2020; @andersen_association_2021]. Such studies cannot assess the risks to immunosuppressed individuals in the population.  Other reports have evaluated the theoretical possibilities for shielding but without empirical data on its effectiveness.  A previous report [@jani_comparison_2020] based on the population of Glasgow did not report age-adjusted rate ratios.

@@

Most previous reports have focused on the ECDPC definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].  The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital acquired infection [@arhai_scotland_hospital_2021].

@@

Our findings support the policy of assigning highest priority for vaccination to those with risk conditions eligible for shielding, as these groups have markedly elevated risks of severe COVID-19. For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke an immune response sufficient to be effective  [@chong_comprehensive_2017].  The relation of severe disease in those eligible for shielding to number of adults in the household raises the question of whether priority should be given to vaccinating  household contacts of solid organ transplant recipients and other vulnerable individuals. The Joint Committee on Vaccination and Immunisation did not recommend this in a report published in December 2020, stating that "at this time there are no data on the size of the effect of COVID-19 vaccines on transmission." [@joint_committee_on_vaccination_and_immunisation_joint_2020].  The Committee however recommended vaccination of  health and social care workers who are "at risk of transmitting infection to multiple vulnerable persons or other staff in a health care environment",  implying some expectation that vaccination would reduce transmission.  On this basis, there is an equally strong case for offering vaccination to household contacts of solid organ transplant recipients without waiting for more definitive evidence of the effect of vaccines on transmission. Even with vaccination of  household contacts, this group will require shielding until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.

Policies for control of transmission in the population and for reducing burden on health services and total deaths have focused on population-wide reduction of social contact. However our analysis provides compelling evidence for a substantial contribution of nosocomial transmission to the burden of severe COVID-19 even during the second wave. A report for NHS England by the Healthcare Safety Investigation Branch noted the challenges of controlling nosocomial transmission and recommended that a national strategy for infection prevention and control should be developed [@healthcare_safety_investigation_branch_investigation_2020].  Vaccination of health care workers is likely to reduce nosocomial transmission, and vaccination of all inpatients should also be considered.  We have not attempted to quantify the onward transmission of COVID-19 from discharged patients, but the PHE modelling study estimated that this could have accounted for 5% of cases requiring hospitalisation in England during the first wave [@knight_contribution_2021]. This suggests  more stringent testing before discharge and quarantine post discharge should be considered. More detailed understanding of how recommended IP policies are being operationalised is also needed [@oliver_could_2021].

# Declarations
## Public and Patient Involvement statement:
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives.

@@

## Transparency declaration
PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained. This manuscript has been generated directly from the source data by a reproducible research pipeline.

@@

## Data sharing statement
The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.

>>>>

@@

format.estci <- function(x.ci) {
    x <- gsub(" \\(", " \\(95\\% CI ", as.character(x.ci))
    x <- gsub(", ", " to ", x)
    #x <- gsub("\\)", paste0(", _p_=\\", x.pval, "\\)"), x)
    #x <- gsub("times", "\\\\times", x)
    return(x)
}

@@

Word counts: abstract 350 words, main text 5771 words.  
2 figures, 4 tables, 26 references

@@

# Abstract  
* _Background_ -  Clinically vulnerable individuals have been advised to shield themselves during the COVID-19 epidemic.  The objectives of this study were to investigate: (1) the risk of severe COVID-19  in  those eligible for shielding, and (2) the relation of severe COVID-19 to transmission-related factors in those in shielding and the general  population.  
* _Methods_ -  All `r nrow(cc.all[CASE==1])` diagnosed cases of COVID-19 in Scotland from 1 March 2020 to `r dateformat(lastdate.specimen)` were matched for age, sex and primary care practice to `r nrow(cc.all[CASE==0])` controls from the general population.  This dataset (REACT-SCOT) was  linked to the list of `r nrow(shielded.full)` individuals identified as eligible for shielding.  Severe COVID-19 was defined as cases that entered critical care or were fatal.  
* _Results_ - With those without risk conditions as reference category, the univariate rate ratio for severe COVID-19 was `r format.estci(table.listedgr3.severecases.nocare["Moderate risk condition", 3])` in those with moderate risk conditions and 
`r format.estci(table.listedgr3.severecases.nocare["Eligible for shielding", 3])` 
in those eligible for shielding.  The highest rate was in solid organ transplant recipients: rate ratio `r format.estci(table.shielded.severecases.nocare["Solid organ transplant", 3])`.  Risk of severe COVID-19 increased with the number of adults but decreased with the number of school-age children in the household. Severe COVID-19 was strongly associated with recent exposure to hospital (defined as 5 to 14 days before presentation date): rate ratio `r format.estci(table.shielded.severecases.nocare["Recent hospital visit/stay", 3])`  overall.   To test for causality, a case-crossover analysis was undertaken; with less recent exposure only (15 to 24 days before first testing positive) as reference category, the rate ratio associated with recent exposure only was `r format.estci(table.tw["Recent.TW.only", 4])`. The population attributable risk fraction for recent exposure to hospital peaked at  `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-05-01"), popattr.frac])`% in May 2020 and again at `r round(100 * coeffs.hosp.timewindow[date==as.Date("2020-12-01"), popattr.frac])`%  in December 2020.  
* _Conclusions_ -  The effectiveness of shielding vulnerable individuals was limited by the inability to control transmission in hospital and from other adults in the household.  For solid organ transplant recipients, in whom the efficacy of vaccines is uncertain, these results support a policy of offering vaccination to household contacts. Mitigating the impact of the epidemic requires control of nosocomial transmission.

\newpage
\linenumbers
<!---
# Author Summary
## Why was this study done?  
* Individuals designated as "clinically extremely vulnerable" -- about 3% of the population -- have been advised to shield themselves during the COVID-19 epidemic

@@

## What did the researchers do and find?  
* Despite the provision of advice and support, the risk of severe COVID-19 remained far higher in those eligible for shielding than in those without risk conditions.

* In both those eligible for shielding and in the rest of the population, the risk of severe COVID-19 was higher in those who were sharing a household with other adults, and was far higher in the 5 to 14 days after exposure to hospital than in those without recent exposure to hospital.

* In the second wave of the epidemic in December 2020, more than half of the cases of severe COVID-19 in the population were attributable to recent hospital exposure.

## What do these findings mean?  
* For vulnerable individuals to be shielded, the risk of hospital-acquired infection must be reduced and support must be provided for other adults in the household to co-isolate with the vulnerable individual.

@@

* Limiting the burden of COVID-19 requires control of hospital-acquired infection. 
--->

@@

# Methods
We linked a national dataset of those eligible for shielding to a case-control dataset (REACT-SCOT) established in Public Health Scotland (PHS) that includes all COVID-19 cases in Scotland and matched population controls.[@mckeigue_rapid_2020]   This case-control dataset is refreshed every few weeks and linked to health records that are used to assign a list of designated "moderate risk conditions". Thus all cases and controls were classified into three categories: no risk condition, moderate risk condition only, eligible for shielding.  The rate ratio of COVID-19 associated with each risk category was estimated  as the conditional odds ratio as described below.

## Ascertainment of eligibility for shielding 
The list of those eligible for shielding was generated by Public Health Scotland from March 2020 onwards by querying health-care information systems including hospital discharge records, prescription encashments, regional cancer chemotherapy databases, blood and transplant registries, for a designated list of diseases and conditions, supplemented direct requests to clinicians in relevant specialties [@health_protection_scotland_search_2020].  The categories designated as eligible for shielding were as listed below [@health_protection_scotland_search_2020; @kanani_caring_2020]:

1. Solid organ transplant recipient
2. Cancer of the blood or bone marrow at any stage of treatment, or people with cancer receiving treatments that affect the immune system 
3. Severe respiratory conditions including cystic fibrosis, severe asthma and severe chronic obstructive airway disease, on home oxygen, severe bronchiectasis, pulmonary hypertension)
4. Rare diseases that increase the risk of infections such as severe combined immunodeficiency, homozygous sickle-cell disease. 
5. People on immunosuppression therapies sufficient to increase risk of infection
6. Pregnant with heart disease
7. Additional conditions, including people on renal dialysis, those who had had a splenectomy  and others identified by clinicians as requiring shielding advice.

@@

## Transmission-related risk factors 
As addresses in the CHI database have been mapped to Unique Property Reference Numbers it was possible to calculate the numbers of adults and children in each household and to augment the coding of care home residence in the CHI database.  Care home residence was assigned using the field in the CHI database, augmented by coding as care home residents the `r care.home.reclassified[1]` individuals aged over 70 in households with 10 or more adults to give a total of `r cc.all[care.home=="Care/nursing home", .N]` care home residents out of the `r cc.all[, .N]` cases and controls.  Linkage to occupational status for health-care workers and teachers was undertaken as described elsewhere [@shah_risk_2020; @mcallister_report_2020].

We used the Scottish Morbidity Records SMR01 (inpatients and day cases) and SMR00 (outpatient attendance) together with the RAPID database to derive variables encoding recent exposure to hospitals. We defined the variable "recent hospital exposure" as any hospital in-patient stay, day case attendance or face-to-face out-patient consultation from 5 to 14 days before presentation date. Restriction of hospital exposure to this time window was intended to exclude consultations caused by COVID-19 symptoms for which testing was delayed by a few days, but to include those exposed to health care facilities in the time interval during which the infection was likely to have been acquired.  Our intention was to capture all relevant exposure in cases and controls, rather than to assign cases as "health-care associated COVID-19" as specified by the European Centre for Disease Prevention and Control (ECDPC) [@european_centre_for_disease_prevention_and_control_surveillance_2020].  The ECDPC case source definitions are restricted to cases who are already hospital in-patients when they first test positive: all other cases are classified as community onset.  <!---Cases admitted 1 to 2 days before first  developing symptoms are classified as community-associated, 3 to 7 days before as indeterminate association, 8 to 14 days before as probable healthcare-associated, and 15 or more days before as definite healthcare-associated. --->

## Statistical methods
The analyses presented here focus on severe COVID-19, defined as the main outcome measure of  of the REACT-SCOT study at the design stage.  Ascertainment of this outcome does not depend upon the extent of testing for SARS-CoV-2, which varied markedly over the study period and is likely to differ between those eligible for shielding and those ineligible.  For the case-control analyses, cases and controls were censored at date of first vaccination and care home residents were excluded, as shielding advice intended for individuals in private households would not have been relevant to care home residents.

@@

To plot the time course of the rate ratios associated with eligibility for shielding and with hospital exposure, these rate ratios were estimated over `r winsize`-day sliding windows of calendar time.  The population attributable risk fraction (PARF) of severe cases for hospital exposure was calculated in each 21-day time window from Miettinen's formula as \ensuremath{p_c \left( r - 1 \right) / r}.  where \ensuremath{p_c} is the frequency of exposure in cases and \ensuremath{r} is the rate ratio [@hsieh_effect_1988].  <!---Results were similar with use of Levin's formula \ensuremath{p \left( r - 1 \right) / \left[ 1 + p \left( r - 1 \right) \right]}  where \ensuremath{p} is the frequency of exposure in controls.--->

@@

To test for causality of the association of recent exposure to hospital with severe COVID-19, we estimated rate ratios associated with recent hospital exposure (5 to 14 days before first testing positive) with less recent exposure (15 to 24 days before first testing positive) as reference category. This analysis was restricted to test-positive cases and their matched controls, so that it does not rely on the imputed presentation dates that were assigned to test-negative cases. The classic case-crossover design compares in cases only the frequencies of exposure in recent and less recent time windows, and estimates the conditional odds ratio  as the ratio of number of cases with recent exposure only to number of cases with less recent exposure only. This can be viewed as a matched-pairs case-control study in which the case and control are the same person in recent and less recent time windows [@maclure_should_2000]. The analysis that we report is a refinement that takes advantage of the availability of a matched control group to contrast the conditional odds of being a case between those with recent exposure only and those with less recent exposure only. This controls for any difference in the population frequencies of exposure or disease incidence between recent and less recent time windows.

@@

In a _post hoc_ analysis we found that the association of recent exposure to hospital with severe disease was largely driven by inpatient exposure [rate ratio `r format.estcipv(table.rapid.coeffs["rapid.recent", 3], table.rapid.coeffs["rapid.recent", 4])`]; the  rate ratios associated with day case exposure [`r format.estcipv(table.daycase.coeffs["daycase.recent", 3], table.daycase.coeffs["daycase.recent", 4])`] or outpatient exposure  [`r format.estcipv(table.opd.coeffs["opd.recent", 3], table.opd.coeffs["opd.recent", 4])`]  were  much lower.   To test if the association of severe COVID-19 with recent inpatient exposure was likely to simply reflect  confounding by time invariant comorbidity, the association was examined by time window as described in the Methods section. Table \ref{tab:timewindow} shows that the rate ratio for severe COVID-19 associated with inpatient exposure only in the recent interval (days 5 to 14 before testing positive), with exposure only in the less recent interval (days 15 to 24 before testing positive) as reference category was `r format.estcipv(table.tw["Recent.TW.only", 4],table.tw["Recent.TW.only", 5])`.   
<!--- tabulate underlying cause of death on death certs of fatal cases
paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(hosp.recent, covid_ucod)))
with(cc.severe[CASE==1], table(hosp.recent, inhosp))
paste.rowpercent(with(cc.all[testpositive.case & deathwithin28==1], table(occup, covid_ucod)))
--->
The frequency of recent inpatient exposure over the period of study in cases not resident in care homes was `r round(100 * p.rapid)`%.  From this and the  rate ratio of `r round(r.rapid, 1)` given above the PARF for recent hospital inpatient exposure  can be calculated from Miettinen's formula as `r round(100 * parf.rapid)`%.  As Table \ref{tab:hcai_rapid}, shows, most of the difference in inpatient exposure classification of severe cases between our definition and the ECDPC definition arises because cases with onset after hospital discharge  would be classified as community onset by the ECDPC classification:  only `r x <- with(cc.severe[CASE==1 & rapid.recent], table(hcai, rapid.recent)); round(100 * sum(x[4:5]) / sum(x))`% of severe cases with inpatient exposure 5-14 days before COVID-19 presentation would be defined as probable or definite hospital-acquired infection (in hospital for at least 8 days before first testing positive) by the ECPDC classification.  The rate ratio for severe disease  associated with probable or definite hospital-acquired infection as defined by the ECDPC classification was `r round(r.hcai, 1)`.

@@

### Risk to those eligible for shielding
In  comparison with those with no risk condition, the highest rate ratio for severe COVID-19 was the `r round(as.numeric(gsub(" .*", "", table.shielded.severecases.nocare["Solid organ transplant", 3])))`-fold rate ratio in solid organ transplant recipients. For other conditions deemed eligible for shielding, the rate ratios were between `r round(min(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[4:8, 3]))))` and `r round(max(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[4:8, 3]))))`, compared to the rate ratio of about `r round(as.numeric(gsub(" .+", "", table.shielded.severecases.nocare[2, 3])))` associated with conditions designated as moderate risk, which include heart disease, diabetes, chronic kidney disease and disabling neurological conditions.  The numbers of pregnant women with heart disease were too small for the risk in this group to be estimated. Other studies of outcome of COVID-19 in solid organ transplant recipients and other individuals using immunosuppressants have been based only on patients admitted to hospital [@avery_inpatient_2020; @andersen_association_2021]. Such studies cannot assess the risks to immunosuppressed individuals in the population.

@@

The role of nosocomial transmission was recognized early in the epidemic. Hospital-acquired infection was suspected in 41% of hospitalised cases seen in January 2020 in one centre in Wuhan [@wang_clinical_2020-2].  A commentary in April 2020 noted that "in Lombardy, SARS-CoV-2 became largely a nosocomial infection" [@boccia_what_2020].  Most previous reports have focused on the narrow definition of probable hospital-acquired infection, rather than the association of disease with exposure to hospital [@howdon_probable_2020].  The ARHAI (Antimicrobial Resistance and Healthcare Associated Infection) unit of NHS National Services Scotland recently reported that among those who died within 28 days of testing positive for COVID-19 (including care home residents), 1015 met the  ECDPC definition of possible or probable hospital-acquired infection [@arhai_scotland_hospital_2021].

@@

Our findings support the policy of assigning highest priority for vaccination to those with risk conditions eligible for shielding, as these groups have markedly elevated risks of severe COVID-19. For solid organ transplant recipients -- the group at highest risk -- it is uncertain whether the vaccine will evoke an immune response sufficient to be effective  [@chong_comprehensive_2017].  As there is now evidence that vaccination reduces transmission to household contacts [@public_health_scotland_covid-19_2021], one way to reduce risk in this group would be to vaccinate their household contacts. Even with vaccination of  household contacts, solid organ transplant recipients will require shielding until passive immunisation is available or the risk of transmission has been reduced to a very low level by herd immunity.

Policies for control of transmission in the population and for reducing burden on health services and total deaths have focused on population-wide reduction of social contact. However our analysis provides compelling evidence for a substantial contribution of nosocomial transmission to the burden of severe COVID-19 even during the second wave. A report for NHS England by the Healthcare Safety Investigation Branch noted the challenges of controlling nosocomial transmission and recommended that a national strategy for infection prevention and control should be developed [@healthcare_safety_investigation_branch_investigation_2020].  Vaccination of health care workers is likely to reduce nosocomial transmission, but vaccination of those booked for elective procedures should also be considered.  We have not attempted to quantify the onward transmission of COVID-19 from discharged patients, but the PHE modelling study estimated that this could have accounted for 5% of cases requiring hospitalisation in England during the first wave [@knight_contribution_2021]. This suggests  more stringent testing before discharge and quarantine post discharge should be considered. More detailed understanding of how recommended infection control policies are being operationalised is also needed [@oliver_could_2021].

# Declarations
<!---## Public and Patient Involvement statement:
This study was conducted under approvals from the [Public Benefit and Privacy Panel for Health and Social Care](https://www.informationgovernance.scot.nhs.uk/pbpphsc/) which includes public and patient representatives. --->

@@

<!---## Transparency declaration
PM, as the manuscript’s guarantor, affirms: that the manuscript is an honest, accurate, and transparent account of the study being reported; that no important aspects of the study have been omitted; and that any discrepancies from the study as originally planned and registered have been explained. This manuscript has been generated directly from the source data by a reproducible research pipeline. --->

@@

## Data Availability
The component datasets used here are available via the Public Benefits Privacy Panel for Health at https://www.informationgovernance.scot.nhs.uk/pbpphsc/ for researchers who meet the criteria for access to confidential data. All source code used for derivation of variables, statistical analysis and generation of this manuscript is available on  https://github.com/pmckeigue/covid-scotland_public.
